---
title: Orchestrating Microbiome Analysis with Bioconductor
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    # These settings below enable word calculation
    filters:
      - at: pre-quarto
        path: _extensions/andrewheiss/wordcount/citeproc.lua
      - at: pre-quarto
        path: _extensions/andrewheiss/wordcount/wordcount.lua
author:
  - name: Tuomas Borman
    affiliations:
      - name: University of Turku
        id: 1
        orcid: 0000-0002-8563-8884
        department: Department of Computing
        address:
        city: Turku
        postal-code: 20014
    email: tvborm@utu.fi
    attributes:
      corresponding: true
  - name: Giulio Benedetti
    affiliations:
        id: 1
        orcid: 0000-0002-8732-7692
    attributes:
      equal-contributor: true
  - name: Geraldson Muluh
    affiliations:
        id: 1
    attributes:
      equal-contributor: true
  - name: Aura Raulo
    affiliations:
        id: 1
  - name: Benjamin Valderamma
    affiliations:
      - name: Organization XK
        id: 2
        department: Department of Y
        address:
        city: City
        postal-code: 0000000
  - name: Artur Sannikov
    affiliations:
      - id: 1
  - name: Stefanie Peschel
    affiliations:
      - name: Organization D
        id: 4
        department:
        address:
        city:
        postal-code:
    email:
  - name: Yihan Liu
    affiliations:
      - name: Cornell
        id: 5
        department: Department of Y
        address: Address
        city: City
        postal-code: 0000000
  - name: OMA consortium
    affiliations:
      - id: 6
  - name: Katariina Pärnänen
    affiliations:
      - id: 1
      - name: University of Helsinki
        id: 6
        department: Department of XYZ
        address:
        city: Helsinki
        postal-code: 20014
  - name: Christian L. Müller
    affiliations:
        - id: 4
  - name: Aki Havulinna
    affiliations:
        - id: 1
  - name: Sudarshan Shetty
  - name: Marcel Ramos
    affiliations:
      - name: Organization X
        id: 9
        department: Department of Y
        address: Address
        city: City
        postal-code: 0000000
  - name: Domenick J. Braccia
    affiliations:
      - name: Organization XA
        id: 10
        department: Department of Y
        address: Address
        city: City
        postal-code: 0000000
  - name: Héctor Corrada Bravo
    affiliations:
      - name: Organization XP
        id: 11
        department: Department of YS
        address: Address
        city: City
        postal-code: 0000000
  - name: Felix G.M. Ernst
  - name: Levi Waldron
    affiliations:
      - name: Organization N
        id: 13
        department: Department of Y
        address: Address
        city: City
        postal-code: 0000000
  - name: Thomaz Bastiaanssen
    affiliations:
      - name: Organization XX
        id: 14
        department: Department of Y
        address:
        city: City
        postal-code: 0000000
    attributes:
      equal-contributor: true
  - name: Himel Mallick
    affiliations:
        id: 5
    attributes:
      equal-contributor: true
  - name: Leo Lahti
    affiliations:
        id: 1
        orcid: 0000-0001-5537-637X
    email: lemila@utu.fi
    attributes:
      corresponding: true
abstract: |
  The expansion of microbiome research has led to the accumulation of
  interlinked datasets encompassing versatile taxonomic and functional
  assays. While critical to advance the field, the analysis of increasingly
  large and heterogeneous multi-modal microbiome data would benefit from
  unified approaches supporting interoperability between methods
  and the design of modular data science workflows. The Bioconductor project
  has recently developed an optimized statistical programming framework for
  multi-assay data integration in computational life sciences. Building on this
  foundation, we introduce a community-developed open-source ecosystem
  for microbiome data science, dedicated to supporting joint analysis
  of hierarchical, interlinked, and heterogeneous datasets that
  are increasingly common in modern microbiome research. This
  interoperable infrastructure comprises of open-source datasets,
  methods, tutorials and an active online community. Its functionality
  and usage are detailed in the OMA book
  (https://microbiome.github.io/OMA/docs/devel), which offers guidance for
  prospective users and contributors.

keywords: [microbiome, Bioconductor, data science, data integration]
bibliography: bibliography.bib
---

<!---
https://communities.springernature.com/posts/how-to-write-an-abstract
-->

# Introduction {#sec-intro}

Modern data science applications critically rely on open-source methods created
by the research community [@Barker2022 ; @Pearson2025] and open
software ecosystems have become central innovation hubs, driving
technological and cultural transformation [@Hocquet2024]. The
Bioconductor project has been established as a major distribution
channel for open data science methods in the life sciences. Nowadays,
its vast developer community maintains over 2,300 quality-controlled R
packages for various fields of life science informatics [@gentleman2004].
Additionally, Bioconductor has evolved into a global community that supports
data science collaboration and training [@drnevich2025] in general, and now the
international calls for enhancing microbiology education in particular
[@Timmis2024].

Microbiome research focuses on the study of microbial communities and
their diversity, function, and interactions with the host and
environment [@moreno-indias2021]. The technologies used in
this field generate vast amounts of high-dimensional, hierarchically structured
data. The expansion of microbiome research has led to the accumulation of
interlinked, multi-modal datasets encompassing versatile taxonomic and
functional assays. Microbiome data scientists are thus increasingly facing
the need to integrate heterogeneous data across multiple modalities while
implementing reproducible workflows, which presents many practical challenges.
First, the unique statistical properties and complexity of microbiome data
necessitate specialized approaches [@moreno-indias2021; @Jeganathan2021;
@Willis2019]. Secondly, the diversity of proposed solutions can make it
difficult to identify the optimal methods and build interoperable workflows
[@Wen2023 ; @Shetty2019]. Thus, the need for standardized microbiome data
science infrastructures  has been widely recognized [@moreno-indias2021]
and a variety of open microbiome data science
frameworks have been developed, including Anvi'o [@Eren2021],
Mothur [@schloss2009], QIIME2 [@bolyen2019], and phyloseq [@mcmurdie2012].
However, platforms that exclusively focus on microbiome data science may not be
able to optimally address the need to link data across different multiple
modalities originating from different research domains.

The Bioconductor project embeds the methods of microbiome data science
into a wider statistical programming framework
alongside with the concurrent development of interoperable methods and
workflows for transcriptomics [@righelli2022], single-cell [@amezquita2020],
metabolomics [@gatto2025], proteomics [@gatto2025; @Anwar2025], and other
fields. Here, we introduce a community-developed open-source framework,
supporting multi-modal data integration in microbiome research. We have
extensively tested and documented the framework to facilitate the dissemination
of evidence-based best practices in microbiome data science.

# Data infrastructure {#sec-infra}

Data scientists routinely
deal with interlinked assays, side information, and hierarchical
data such as ontologies, trees, or networks. The ability to integrate
heterogeneous data elements is essential in biological research, where the
individual components of a system can be seldom studied in isolation.
The challenge for a data scientist is that different methods often require
variable and inconsistent input data formats, leading to technical
complications and time lost on repeated data wrangling.
Bioconductor addresses this through standardized data structures supported by
an ecosystem of interoperable methods, allowing more time to be invested in core
analytical tasks.

Bioconductor data structures support statistical analysis of complex
data combinations [@huber2015] and are widely adopted by various
user communities. They have been implemented through specialized
object-oriented classes that integrate multiple data elements into
a single structure known as a *data container*
[@chambers2014]. The SummarizedExperiment (SE) class represents
the primary data container underlying the Bioconductor ecosystem
[@huber2015; @morgan2025]. Its position among the top-1\% most
downloaded packages in Bioconductor reflects its widespread recognition and
increasing adoption among developers (@fig-bioc_packages). SE provides a
standardized solution to store tabular data by linking numeric abundance
matrices with side information on features (rows) and samples (columns).
Side information can include secondary details
on the experiment, such as sample origin, collection time, or health status.

The interoperability of SE with other Bioconductor data structures
facilitates the transfer of methods across different fields of omics research.
This represents a major strength, which has facilitated the development of
modular and scalable data science methods [@drnevich2025].
The shared data structures support the transfer of methods between
omics fields, yielding improved overall quality control and user support.
Consequently, the SE-based framework has been extended to
similar data integration tasks, notably in single-cell
and spatial omics through the SingleCellExperiment (SCE) [@amezquita2020] and
SpatialExperiment [@righelli2022] classes, respectively.

```{r}
#| label: fig-bioc_packages
#| eval: true
#| echo: false
#| message: false
#| out-width: 70%
#| fig-cap: "**Adoption of the SummarizedExperiment (SE) data container among Bioconductor developers.** The growth in the number of Bioconductor packages supporting SE over time is shown, categorized by their respective fields. For some of the packages SE support has been added after their original release."

if( !require("BiocManager") ){
    install("BiocManager")
}
pkgs <- c("BiocPkgTools", "lubridate", "tidyverse", "see")
temp <- sapply(pkgs, function(pkg){
    if( !require(pkg, character.only = TRUE) ){
        BiocManager::install(pkg)
        library(pkg, character.only = TRUE)
    }
})

# Get packages and info based on their BiocViews
pkgs <- biocPkgList()
#
pkgs[["SE"]] <- sapply(pkgs[["Depends"]], function(x) any(grepl("summarizedexperiment", x, ignore.case = TRUE)) )
pkgs <- pkgs[pkgs[["SE"]], ]
fields <- sapply(pkgs$biocViews, function(x){
    field <- "Other"
    if( any(grepl("genomic|genetic|gene|genom|DNA", x, ignore.case = TRUE)) ){
        field <- "Genomics"
    }
    if( any(grepl("proteomics|protein", x, ignore.case = TRUE)) ){
        field <- "Proteomics"
    }
    if( any(grepl("metabolomics|metabolome|metabolite|Lipidom|massspectro", x, ignore.case = TRUE)) ){
        field <- "Metabolomics"
    }
    if( any(grepl("transcripto|RNA-seq|RNA", x, ignore.case = TRUE)) ){
        field <- "Transcriptomics"
    }
    if( any(grepl("immuno", x, ignore.case = TRUE)) ){
        field <- "Immunology"
    }
    if( any(grepl("cytom", x, ignore.case = TRUE)) ){
        field <- "Cytometrics"
    }
    if( any(grepl("microarray|chip", x, ignore.case = TRUE)) ){
        field <- "Microarray"
    }
    if( any(grepl("single-cell|singlecell", x, ignore.case = TRUE)) ){
        field <- "Single-cell"
    }
    if( any(grepl("metagenom|microbiome|16S|microbiota|amplicon|shotgun|microb|metatranscript|metametabolo|metaproteo", x, ignore.case = TRUE)) ){
        field <- "Microbiome"
    }
    return(field)
})
pkgs[["Field"]] <- fields
# Get download stats
df <- biocDownloadStats()
# Subset to include only packages which BiocView we have
df <- df[df[["Package"]] %in% pkgs$Package, ]
# Add field info
df[["Field"]] <- pkgs[match(df[["Package"]], pkgs[["Package"]]), "Field"][[1]]
# df <- df[df$Package %in% pkgs$Package, ] # Take into account only current packages
pkgs <- pkgs[, c("Package", "SE", "Field")]
# Get table that shows which package is available in each date
dates <- df$Date |> unique()
dates <- dates[ dates < floor_date(Sys.time(), "month") ]
available <- lapply(dates, function(date){
    temp <- df[df[["Date"]] == date, ]
    temp <- temp[match(unique(df$Package), temp$Package), ]
    temp <- temp[["Nb_of_distinct_IPs"]] > 0 | temp[["Nb_of_downloads"]] > 0
    temp[is.na(temp)] <- FALSE
    temp <- as.numeric(temp)
    return(temp)
})
available <- do.call(cbind, available) |> as.data.frame()
rownames(available) <- unique(df[["Package"]])
colnames(available) <- dates
#
ind <- pkgs[match(unique(df[["Package"]]), pkgs[["Package"]]), "Field"][[1]]
pkgs_date <- rowsum(available, group = ind)
# Put into long format
pkgs_date <- pkgs_date |>
    rownames_to_column("Field") |>
    pivot_longer(
        cols = colnames(pkgs_date)[ !colnames(pkgs_date) %in% c("Field")],
        names_to = "Date",
        values_to = "N"
    )
pkgs_date[["Date"]] <- pkgs_date[["Date"]] |> as.Date()

# Create a plot that shows number of package through date
p <- ggplot(pkgs_date, aes(x = Date, y = N, fill = Field)) +
    geom_area() + theme_classic() + scale_fill_okabeito() +
    labs(x = "Year", y = "Number of packages") +
    scale_y_continuous(expand = c(0, 0), limits = c(0, NA))
p
```

**Hierarchical data structures.** Microbiome data scientists frequently need to
integrate information on feature phylogenies and sample hierarchies
in order to accurately reflect fine-scale variability in microbiome data.
While maintaining interoperability with
the broader Bioconductor ecosystem, the TreeSummarizedExperiment
(TreeSE) class [@huang2021] extends the SE and SCE data structures to
hierarchical datasets by incorporating feature and sample organisations
as tree structures (@fig-datacontainers a). Moreover, TreeSE supports the
storage of other common aspects of microbiome data, for example reference 
sequences, and incorporates results from common analytical procedures such as
dimensionality reduction, statistical transformations, and agglomeration by 
taxonomic, functional or other information. The interoperability with other 
Bioconductor classes facilitates scalable analysis and the design of modular 
yet flexible workflows. The direct interoperability with the widely adopted 
SE data science ecosystem distinguishes our TreeSE-based approach from phyloseq, 
which is another popular Bioconductor data container originally 
developed with a focus on 16S amplicon data analysis [@mcmurdie2012]. 
TreeSE is a superset that can accommodate all elements of a phyloseq object,
while also offering improved support for multi-modal data and often substantial
gains in memory efficiency and computing speed (@fig-benchmarking). Overall, the
TreeSE approach outperforms phyloseq in these benchmarks. While speedyseq is
faster in certain operations, TreeSE scales more efficiently in both execution
time (t) and allocated memory (m) as sample size (n) increases.  Conversion
functions and detailed cheat sheets between the two formats are available to
bridge the gap between the respective user communities.

![***Bioconductor data containers for microbiome data.*** **(a)
TreeSummarizedExperiment (TreeSE).** TreeSE supports hierarchical and
multi-table analysis in microbiome data science [@huang2021]. It links the
numeric abundance matrix and potential alternative versions (statistical
transformations, different taxonomic levels, dimensionality reduction) with
tabular and hierarchical side information on the features (rows) and samples
(columns). Additional slots for reference sequences and experiment metadata are
available. (Figure modified from [@huang2021]) **(b) MultiAssayExperiment
(MAE).** MAE facilitates integration of data from multiple experiments
[@ramos2017]. (Figure from
[@ramos2017])](figures/datacontainers.png){#fig-datacontainers}

![**Comparison of execution time and memory consumption between alternative
methods in Bioconductor** TreeSE, phyloseq and speedyseq were benchmarked for
five common analytical operations applied to random subsets of samples from a
large study on wild baboons [@Grieneisen2021] (accessible through the
microbiomeDataSets Bioconductor package). Execution time (s) values and error
bars represent the mean and standard error (SE) over 10 iterations,
respectively. Allocated memory (MB) was estimated based on a single iteration.
Overall, TreeSE outperformed phyloseq. While speedyseq achieved higher
performance in the supported operations, TreeSE scaled more efficiently in both
execution time and memory allocation with increasing sample size (n). The
benchmarking analysis was conducted in R using the bench library with 8 CPU
cores and 32 GB RAM. All operations, except for UniFrac dissimilarity with
phyloseq, could also be performed on a regular laptop (*e.g.* 4 CPU cores and
16 GB RAM). The source code for the benchmarking experiments can be accessed
through the OMA online book.](figures/benchmarking.png){#fig-benchmarking}

**Multi-assay data structures.** When integrating data across multiple
modalities, such as taxonomy, metabolomics, or host genomics, more flexible
sample mapping may become necessary. Multi-omics data is often sparse, and
samples can be only partially matched between experiments. In other cases,
a single sample in one modality can be matched by two or more samples in
another modality [see @Husso2023]. The MultiAssayExperiment (MAE) data
container [@ramos2017] facilitates the integration of multiple data objects
representing different omics modalities (see @fig-datacontainers b).
Whereas TreeSE is
restricted to a one-to-one sample mapping between different data tables, MAE
supports flexible mappings between multiple experiments from the same study.
Integration of multi-table data without such a structured framework often
leads to ambiguous mappings and inconsistent results.
For instance, diagonal integration methods
[@xu2022; @argelaguet2021computational] &ndash;
where shared features across data types may not be explicitly defined &ndash;
can benefit substantially from such a systematic framework
that enforces anchor alignment. This mirrors lessons from single-cell
data integration, where shared anchors and hierarchical mapping
have proven essential for interpretability. Flexible linking of samples across
several  data objects supports efficient data integration, methods sharing, and
overall interoperability.

@tbl-containers summarizes the microbiome-related data containers, while
@tbl-elements introduces the available data elements in each container.

| Data container              | Target use                                                       |
|-----------------------------|------------------------------------------------------------------|
| SummarizedExperiment (SE)[@huber2015]     | generic data container for experiments with multiple assays      |
| TreeSummarizedExperiment (TreeSE)[@huang2021] | tailored data container for hierarchically organized experiments |
| MultiAssayExperiment (MAE)[@ramos2017]    | tailored data container for sets of multi-omic experiments       |

: Recommended data containers for microbiome analysis. {#tbl-containers tbl-colwidths="[50,50]"}

| Slot          | Content                                      | SE | TreeSE | MAE$^a$ |
|---------------|----------------------------------------------|:--:|:------:|:-------:|
| **Tables**    |                                              |    |        |         |
| assays        | tables of features (rows) and samples (cols) | X  | X      | X       |
| altExps$^b$   | experiments with variable number of features |    | X      | X       |
| reducedDims   | results of dimensionality reduction          |    | X      | X       |
| **Rows**      |                                              |    |        |         |
| rowData       | side information on features                 | X  | X      | X       |
| rowPairs      | linkage between pairs of associated features |    | X      | X       |
| rowTrees      | hierarchical organization of features        |    | X      | X       |
| referenceSeq  | reference sequences of features              |    | X      | X       |
| **Columns**   |                                              |    |        |         |
| colData       | side information on samples                  | X  | X      | X       |
| colTrees      | hierarchical organization of samples         |    | X      | X       |
| sampleMap$^c$ | flexible sample mapping across experiments   |    |        | X       |
| **Other**     |                                              |    |        |         |
| metadata      | additional information on the experiment     | X  | X      | X       |

: Key data elements in the employed data containers. {#tbl-elements tbl-colwidths="[17,63,3,9,8]"}

$^a$ While MAE does not contain conventional slots (with the exception
of colData), its experiments can include any number of data containers that
provide those slots. $^b$ altExp samples must match assay samples in a 1-to-1 mapping.
$^c$ sampleMap supports diverse mappings between samples from different experiments
(1-to-1, 1-to-many, many-to-1 and many-to-many).

**Data resources.** Several open microbiome data resources are supported, enabling
direct data import into the aforementioned data containers for further downstream
analysis within the  Bioconductor methods ecosystem (see @tbl-dataresources).
Examples of the available data resources include
curatedMetagenomicData [@pasolli2017], HoloFood [@rogers2024],
microbiomeDataSets [@microbiomedatasets], and the EBI/MGnify database
[@gurbich2023]. In addition, Bioconductor packages often
include built-in demonstration datasets. Collectively, these resources
encompass taxonomic and functional profiles from thousands of
published microbiome studies and tens of thousands samples that can
be accessed with the available tools for research and educational purposes.

| Data resource                                        | Package                       | # studies or datasets | # samples |
|-------------------------------------------------|------------------------------------------|-----------------------|-----------|
| Curated metagenomic data [@pasolli2017] | curatedMetagenomicData [@pasolli2017]    | 93                    | 22,588    |
| HoloFood [@rogers2024]                          | HoloFoodR [@holofoodr]                   |                      - | 9,990     |
| MGnify [@gurbich2023]                           | MGnifyR [@mgnifyr]                       | 5129                  | 616,138   |
| microbiomeDataSets [@microbiomedatasets]        | microbiomeDataSets [@microbiomedatasets] | 6                     | 19,100    |

: Open microbiome data resources supporting the framework. {#tbl-dataresources tbl-colwidths="[30,40,15,15]"}

# Data preparation {#sec-process}

While microbiome data scientists continue to develop
specialized solutions [@Jeganathan2021; @Willis2019;
@MarcosZambrano2023], the implementations typically differ in their
expected input and output formats. Common data standards support
modular data science workflows, speeding up methods development,
benchmarking, and replication. Here, we describe an ecosystem of
software packages that has recently emerged to support microbiome data
integration and analysis based on the data containers, (Tree)SE and MAE
(see @fig-workflow).

The framework focuses primarily on the downstream analysis of
high-throughput sequencing data from microbiome experiments after the
original measurements have already been summarized into abundance
tables representing, for example, taxonomic or functional profiles. Taxonomic
abundance tables are one of the most commonly encountered data types.
Common measurement platforms include amplicon-based marker gene
studies (*e.g.* 16S rRNA) and metagenomic sequencing, but alternative profiling
techniques are available including phylogenetic microarrays
[@RajilicStojanovic2009] and high-density optical mapping
[@Abakumov2024]. Microbiome analyses can also yield variable metagenomic,
and functional features. The interoperability of the data
formats supports modular workflow design for integrating different
omics through the Bioconductor ecosystem. This
decisively mitigates reproducibility challenges posed by rapidly
evolving reference databases and profiling tools
(*e.g.* MetaPhlAn2 vs. MetaPhlAn4) by enabling structured coexistence and
direct comparison within a unified, version-resilient framework
[@blanco2023; @truong2015].

**Data import.** Most users will need to import data for their own
analyses. The TreeSE data container can be populated with microbiome
data from common tabular data types but non-standard file formats may
require additional data wrangling. To streamline data
import, we provide importers for a variety of standard microbiome file
formats including BIOM [@mcdonald2012], MetaPhlAn/HUMAnN (MetaPhlAn
v. 2-4, HUMAnN v. 3) [@mciver2017], Mothur [@schloss2009], QIIME2
[@bolyen2019] and the TAXonomic Profile Aggregation and
STAndardisation format (taxpasta) [@beber2023]. These importers can
automatically load the data into the appropriate data containers, thus
forming a bridge between low-level sequence processing methods and the
high-level statistical programming framework in R. Moreover,
converters between alternative data structures in R are available
supporting seamless conversions between TreeSE and BIOM, DADA2
[@callahan2016] and phyloseq (the mia::convert\* functions), and
TreeSE-based microbiome data objects can be exported into external services
such as MicrobiomeDB [@Oliveira2018] and in other languages
(*e.g.* Julia). These bridges facilitate the broadest possible access to
microbiome analysis methods.

**Quality control and exploration.** After importing the data into R,
the recommended first step involves basic data exploration and quality
control [@zuur2010]. In addition to methods for identifying
contaminant sequences and calculating summary statistics, the
framework includes a versatile set of custom visualization methods to
assess data variability, outliers, homogeneity of variance and other
aspects that may need to be considered in downstream data
analysis. The mia and miaViz packages, for instance, provide
convenient access to many operations for common exploratory tasks in
microbiome analysis and visualization for TreeSE objects
(@fig-downstream_analyses).

**Filtering and agglomeration.** Microbiome datasets may contain tens of
thousands of unique features depending on the sequencing resolution,
databases, and the scope of analysis. To summarize data into biologically
meaningful groups or to reduce multiple comparisons the data are often
subsetted, filtered or agglomerated to higher-level taxonomic or functional
categories or stratified by sample groups.he package ecosystem and the mia
package feature various
tools for such tasks. For instance, the prevalence methods can be
useful to remove very rare features that might be sequencing artifacts
or present otherwise limited value for the analysis. The mechanism of
alternative experiments (altExp) supports data agglomeration
by taxonomic levels and by other features such as
pathogeneicity or functional potential (*e.g.* butyrate production
[@Kullberg2024], gut-brain modules [@VallesColomer2019], and other
host-associated signatures [@geistlinger2024]).
The supporting methods incorporate the agglomerated levels
within the same data object for parallel operations, and typically
also take care of agglomerating the associated tree
information and the alternative assays. The filtering,
subsetting, and agglomeration operations are thus supported in both feature
as well as sample spaces, and simultaneously
propagate to cover the relevant alternative assays, alternative
experiments, ordinations, and side information. This supports the
efficient management of interlinked assays and saves time for dedicated
expert tasks.

**Transformation.** Common statistical transformations in
microbiome research include relative abundance [@gloor2017],
centered log-ratio (CLR) and its robust version (rCLR), which are commonly used
to mitigate compositionality bias [@Martino2019], and the tree-aware
phylogenetic ILR (phILR) transformation [@silverman2017]. Data
rarification is a related concept that has been suggested to control
for uneven sequencing depths and it remains widely debated in
microbial ecology [@mcmurdie2014 ; @schloss2024].
Our framework supports these and a variety of other
transformations commonly applied to microbiome data.

**Data enrichment.** Finally, incorporating additional data is a
frequent operation in applied data analysis, and it often takes place
in an iterative fashion. The data containers
support gradual and organized accumulation of interlinked data elements
including transformations, aggregated data, dimension reduction, and derived
data on samples and features. We generally recommend
that the data objects are initialized at the beginning of the
workflow, where they can then be gradually enriched as the project advances.
By systematically extending the upstream data preparation
step &ndash; instead of ad-hoc calculations scattered along the
workflow &ndash; the data scientist can ensure well-organized and transparent
records of data transformations. Data can be enriched by
importing new external information as well as by deriving new
information from the data itself into the same data container as
described in the next section.

![**Examples of Bioconductor tools along the microbiome data science workflow.** A typical microbiome workflow starts with taxonomic and functional annotation. The processed microbiome data can then be imported to Bioconductor data container (TreeSE), or integrated with other omics using the MAE container for downstream statistical analysis. (Figure adapted from [@amezquita2020])](figures/workflow.png){#fig-workflow}

# Downstream data analysis {#sec-analysis}

The new Bioconductor framework for microbiome data science improves the management of
integrative analysis workflows. It facilitates systematic linking of the different data
elements and analysis outcomes, helping to shift the focus towards multi-modal analyses and
methods development while supporting standard analysis tasks in microbiome research.

Microbiome data analysis often relies on specialized statistical approaches
as data sparsity, zero-inflation, compositionality, and heteroschedasticity
pose challenges for standard data analytical methods
[@gloor2017; @Jeganathan2021]. Moreover, microbiome data is hierarchically
structured across taxonomic, ecological, and spatial levels
[@moreno-indias2021; @Ruuskanen2021]. Alternative processing pipelines
introduce different biases and generate varying data formats
[@quince2017]. These and other unique characteristics have
wide-reaching implications for microbiome analysis, necessating dedicated
analysis methods [@moreno-indias2021].
Our statistical programming framework supports the community-driven development
and benchmarking of the solutions. The following sections provide an
overview of the available tools.

**Community indicators.** Various global indicators are available
to quantify the microbial community state or other sample
properties. Community diversity represents a common measure in
microbial ecology [@bastiaanssen_bugs1_2023]
(@fig-downstream_analyses). Alpha diversity metrics generally quantify
the observed or estimated number of distinct species ("richness") and
their distribution ("evenness"). Many diversity indices (e.g.
Shannon index), combine these two aspects. Phylogenetic diversity
measures additionally consider the phylogenetic relatedness of
community members [@faith1992]; the new implementation of the Stacked Faith's
Phylogenetic Diversity can accelerate the performance by 1-2
orders of magnitude [@Armstrong2021]. Enhanced analyses with
repeated rarefaction are supported [@Schloss2024b] albeit the use of
rarefaction remains controversial [@schloss2024; @Willis2019b]. Other
global indices of community state include measures of dominance,
rarity, skewness, kurtosis, modality, library size (number of reads),
or absolute abundance quantification. Sample information in the TreeSE object
(colData) can be enriched with these derived quantities. By default,
the mia::addAlpha method returns a set of complementary alpha
diversity measures [@Cassol2025].

**Community similarity** is commonly evaluated with beta diversity
measures, which quantify multivariate dissimilarity and can be used to
quantitatively compare community composition. The
framework supports common indices in microbial ecology including
Bray-Curtis, Jaccard, Aitchison [@bastiaanssen_bugs1_2023 ; @Martino2019],
and the phylogeny-aware UniFrac [@lozupone2005]. The community
(dis)similarities are commonly visualized with unsupervised ordination
methods that project the data on a lower-dimensional display
[@goodrich2014]. Unsupervised techniques, Principal Component
Analysis (PCA) and Principal Coordinates Analysis (PCoA) also known as 
Multi-Dimensional Scaling (MDS), and Non-metric MDS (NMDS)
(@fig-downstream_analyses) are common choices, with enhanced support
for microbiome analysis and visualization. Statistical tests
(*e.g.* Permutational Multivariate Analysis of Variance) and supervised
visualization methods, such as the
distance-based redundancy analysis (dbRDA) can complement the
unsupervised analyses, and help to uncover and quantify co-variation
between community composition and covariates. This can benefit from
compositionally robust measures (*e.g.* robust
Aitchison distance [@Martino2019]), or phylogeny-aware dissimilarity
measures specifically developed for microbiome research (*e.g.* Unifrac
[@lozupone2005]). Sample dissimilarities provide the basis for many
specialized microbiome techniques including microbial community typing
(clustering) with Dirichlet Multinomial Mixtures (DMM) [@Holmes2012]
and community state types (CST) [@DiGiulio2015], or factorization into
community signatures [@Frioux2023]. These and other multivariate
methods are provided for instance by mia, miaViz, and
additional single-cell packages (scater, scuttle [@amezquita2020]).

**Differential abundance and mediation analysis.** Differential abundance 
analysis (DAA)
and differential prevalence analysis (DPA) methods are used to
identify microbial taxa whose abundances vary across study groups or
continuous gradients (@fig-downstream_analyses). Many dedicated DAA
methods in microbiome research support the TreeSE framework including methods
such as ANCOM-BC2 [@lin2024], LimROTS [@Anwar2025], MaAsLin3
[@nickols2024], and radEmu [@Clausen2024]. Recent studies by us and
others indicated that elementary methods, for example log-transformed
TSS-counts coupled with a linear model, or presence/absence data with
logistic regression often yield the most consistent results
[@Pelto2025]. The standard DAA typically focuses on individual
features, ignoring interactions and potential co-abundance
variation. Thus, we support co-abundance analyses; agglomerating
collinear features as a preprocessing step can reduce multiple
testing, increase statistical power, and support interpretation but is
also prone to subjective parameterization and limited
generalizability. The grouping can be based on clustering of the
abundance profiles or on external information drawn from phylogenetic
relations, taxonomic classifications or functional properties
(*e.g.* butyrate producers [@Kullberg2024]). Beyond these,
advanced methods—such as causal mediation analysis tailored to microbiome
multi-omics data—are already being developed using SummarizedExperiment
[@jiang2025], highlighting its readiness to support emerging innovations in
microbiome analytics.

**Function- and sequence-level analyses.** Taxonomic analysis is
frequently complemented by functional analyses, either based on
functional predictions derived from amplicon or metagenomic
sequences, or from parallel metabolomic or other functional
assays. Our framework not only supports incorporation of functional
abundance tables within the same data objects but specialized methods
on functional data are available; for instance, regarding metabolome
analyses the notame package [@klavus2020] recently added SE support,
and the R4MS project
provides many interoperable tools [@gatto2025]. The framework has
found recent applications in extended analyses of metagenomic
features, *i.e.* antimicrobial resistance gene composition, diversity,
and load [@Salehi2025]. We have converted a large-scale compilation of
resistome profiles in 8,972 metagenome samples [@Lee2023] in the TreeSE format
for research and educational purposes.

**Spatio-temporal, ecological and epidemiological models.** Microbiome time
series [@berg2020] encompass a wide range of study designs, ranging
from short to long follow-up times, sparse to dense data, univariate
to multivariate and multi-omic monitoring [@Velten2022], and real,
pseudo, or simulated time series. Moreover, an increasing number of
studies have incorporated the spatial dimension
in microbiome analysis across scales [@Ruuskanen2021]. While spatial
and temporal analyses require specialized methods, the miaTime R package
offers dedicated solutions for longitudinal microbiome studies.
In addition, Bioconductor provides methods from other application domains,
among which single-cell sequencing [@amezquita2020] and spatial transcriptomics 
[@righelli2022], which use closely related data containers
(@fig-downstream_analyses). Spatial and temporal information can be
incorporated as a covariate or by estimating
temporally or spatially resolved metrics, for instance divergence,
auto-correlation and modality. Another type of temporal analysis 
includes the time-to-event, or survival models, which are encountered
in, for example, epidemiological and population cohort studies; specialized
survival analysis
strategies for microbiome data have been proposed [@Salosensaari2021;
@Calle2023]. We also provide methods to simulate time series from
standard models of microbial community dynamics including the
Hubbell's neutral model, generalized Lotka-Volterra, and
consumer-resource model through the miaSim package [@gao2023].

**Multi-modal integration.** An increasing number of studies integrate
microbial abundance data with further omics, including transcriptomics,
proteomics, and host genomics [@nyholm2020; @Qin2022;
@moreno-indias2021; @Hintikka2021] (@fig-downstream_analyses). As
demonstrated above, Bioconductor data containers support combinations
of complex omics data. [@huber2015; @ramos2017; @amezquita2020].
Statistical methods for multi-table analysis in microbiome research
can be broadly categorized into three groups: Machine Learning (ML)
predictive modelling [@mallick2024], association-based modelling, and
latent variable modelling.  [@Jeganathan2021; @moiz2025;
@moreno-indias2021; @mallick2024; @bastiaanssen2023;
@argelaguet2020]. Some of the currently supported multi-omic methods
include cross-correlation analyses, canonical correspondence analysis
(CCA), anansi [@bastiaanssen2023], and joint-RPCA [in press].

**Other methods and programming environments.** A range of other methods is
employed in microbiome
data science. Examples include network analysis [@Peschel2021],
ML-based prediction and classification [@Topcuoglu2021], enrichment
analyses via BugSigDB [@geistlinger2024] or Microbe Set Enrichment
Analysis (MSEA) [@Kou2020]. The recent adoption of of interoperable
data containers and the methodological similarities in related fields
can greatly expand the selection of available toolkits in emerging
areas. In addition to the internal coherence within the Bioconductor ecosystem, 
this framework supports integration with external platforms through interfaces, 
namely `reticulate` for Python, `Rcpp` for C++, and compatibility with
pipelines  bioBakery [@navarro2024; @mciver2017]. These connections empower
users to combine the best tools available across environments and languages
while maintaining structured data interoperability via standard containers like
TreeSE and MAE.

```{r}
#| label: downstream_analyses
#| eval: false
#| echo: false
#| message: false

# Load packages
pkgs <- c("maaslin3", "mia", "miaTime", "microbiomeDataSets")
temp <- sapply(pkgs, function(x)
    suppressMessages(require(x, character.only = TRUE)))

# Load data
tse <- SprockettTHData()
# Subset by taking only fecal samples from infants
tse <- tse[, tse[["Sample_Type"]] %in% c("Feces")]
tse <- tse[, tse[["Age_Class"]] %in% c("Infant")]

# Agglomerate and apply transformations
tse <- agglomerateByRanks(tse)
tse <- transformAssay(tse, method = "relabundance", altexp = altExpNames(tse))
tse <- transformAssay(
    tse, method = "clr", pseudocount = TRUE, altexp = altExpNames(tse))
altExp(tse, "prev_genus") <- subsetByPrevalent(
    altExp(tse, "Genus"), prevalence = 0.5, detection = 0.001)
rownames(altExp(tse, "prev_genus")) <- gsub(
    "/| ", "_", rownames(altExp(tse, "prev_genus")))

# Calculate alpha diversity
tse <- addAlpha(tse, index = "faith")

# Perform PCoA
tse <- addMDS(tse, assay.type = "counts", method = "unifrac", niter = 10L)

# Perform dbRDA
tse <- addRDA(
    tse, assay.type = "clr", method = "euclidean",
    formula = x ~ Age_Months + Country, na.action = na.exclude
    )
# Remove the RDA object as it takes lot of space
attr(reducedDim(tse, "RDA"), "obj") <- NULL

# Perform DAA
res_full <- maaslin3(
    input_data = altExp(tse, "prev_genus"),
    output = "DAA_example",
    formula = "~ Age_Months + Sex + Country",
    normalization = "TSS",
    transform = "LOG",
    verbosity = "ERROR"
)
res <- res_full[["fit_data_abundance"]][["results"]]
res <- res[res[["value"]] == "Age_Months",  ]
signif_feat <- res[
    !is.na(res[["qval_individual"]]) & res[["qval_individual"]] < 0.05,
    "feature"]
rowData(altExp(tse, "prev_genus"))[["signif"]] <- rownames(
    altExp(tse, "prev_genus")) %in% signif_feat

# Calculate time divergence
tse <- addStepwiseDivergence(
    tse, time.col = "Age_Months",
    group = "Subject_ID", method = "bray",
    tree = rowTree(tse)
    )

# Save object so that it can be used in visualization step without running the
# analyses
saveRDS(tse, "data/tse.rds")
```

```{r}
#| label: fig-downstream_analyses
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Common visualization methods applied to time series data on infants from three different countries. **a**, Phylogenetic relationships. **b**, Phyla abundances of the samples ordered by age, stratified by country and gender. Actinobacteria is more present in Bolivia and Bangladesh compared to Finland.  **c-d**, Multidimensional Scaling (MDS) using Unifrac distances and Distance-based Redundancy Analysis (dbRDA) based on Aitchison distance, respectively. Both ordination methods shows clear pattern separating microbial profiles in different countries. dbRDA shows additionally association between microbial profile and age. **e-f**, Faith's phylogenetic diversity by country and age respectively. **g**, Differential Abundance Analysis (DAA) results. Two genera exhibit association with age. **h**, Unifrac divergence between consecutive time points. Results indicate slow stabilization of microbial communities."
#| fig-width: 19
#| fig-height: 10

# Load packages
pkgs <- c("miaViz", "patchwork", "scater")
temp <- sapply(pkgs, function(x)
    suppressMessages(require(x, character.only = TRUE)))

# Load data
file_path <- "data/tse.rds"
if( !file.exists(file_path) ){
    stop("Run manually the chunk above. '", file_path, "' cannot found.")
}
tse <- readRDS(file_path)

# Get phylum names
phylum_names <- sort(unique(rowData(altExp(tse, "Phylum"))[, "Phylum"]))
# Named vector of phylum colors
phylum_colors <- setNames(
    scater:::.get_palette("tableau20")[seq_along(phylum_names)],
    phylum_names)

# Visualize phylogeny
p_tree <- plotRowTree(
    altExp(tse, "Family"), edge.colour.by = "Phylum", layout = "ape") +
    scale_color_manual(values = phylum_colors) + # Manually set colors
    guides(color = "none")

# Visualize abundances with barplot
p_bar <- plotAbundance(
    altExp(tse, "Phylum"), assay.type = "relabundance",
    col.var = c("Country", "Sex"), facet.cols = TRUE, scales = "free",
    order.col.by = "Age_Months") +
    guides(fill = guide_legend(title = "Phylum")) +
    scale_color_manual(values = phylum_colors)

# Visualize alpha diversity
p_alpha <- plotColData(
    tse, y = "faith", x = "Country", colour_by = "Country",
    show_boxplot = TRUE, show_violin = FALSE, point_alpha = 0.2) +
    labs(y = "Faith", x = "Country") +
    theme(legend.position = "none")
p_alpha2 <- plotColData(
    tse, y = "faith", x = "Age_Months") +
    geom_smooth(method = "lm") +
    labs(y = "Faith", x = "Age in months")

# Visualize PCoA
p_mds <- plotMDS(tse, colour_by = "Country") +
    theme(legend.position = "inside",
          legend.justification = c("right", "top"))

# Visualize dbRDA
p_rda <- plotRDA(tse, "RDA", colour.by = "Country", add.significance = FALSE) +
    theme(legend.position = "none") # Already present from p_mds

# Visualize abundances vs age
p_daa <- plotExpression(
    altExp(tse, "prev_genus"),
    features = rownames(altExp(tse, "prev_genus"))[rowData(altExp(tse, "prev_genus"))[["signif"]]],
    assay.type = "relabundance", x = "Age_Months", ncol = 3L,
    ) +
    scale_y_log10() +
    geom_smooth() +
    labs(x = "Age in months", y = "Relative abundance")

# Visualize divergence
tse_sub <- tse[, !is.na(tse[["time_diff"]]) & tse[["time_diff"]]>=0.8 & tse[["time_diff"]]<=1.2]
p_divergence <- plotColData(tse_sub, y = "divergence", x = "Age_Months") +
    geom_smooth(method = "lm") +
    labs(x = "Age in months", y = expression(Delta ~ "UniFrac" ~ "(N, N-1)") ) +
    theme(legend.position = "none")

# Combine plots
p1 <- (p_tree / p_bar) +
    plot_layout(guides = "collect")
p2 <- (p_mds / p_rda) +
    plot_layout(heights = c(1, 1))
p3 <- ((p_alpha|p_alpha2) / p_daa / p_divergence) +
    plot_layout(heights = c(1, 1, 1))
p <- (p1 | p2 | p3) +
    plot_layout(widths = c(0.33, 0.4, 0.32)) +
    plot_annotation(tag_levels = "a") &
    theme(plot.tag = element_text(size = 15, face = "bold"))
p
```

# Accessible analysis and community {#sec-community}

Modern science is embodied by a critical community capable of
assessing discoveries and replicating results [@Wootton2016].
Open software and the social structures around it
present a timely manifestation of this principle, with technical,
legal, sociological and epistemic consequences [@Hocquet2024]. FAIR
principles for software support
access to well-documented and user-friendly methods [@Barker2022],
and are facilitated by the extensive quality control and
documentation standards enforced by Bioconductor [@huber2015].
In addition to these technical
demands, the development and quality control of academic software
critically relies on active critical source communities that can
provide continuous benchmarking and support the identification and
dissemination of evidence-based best practices [@berg2020 ;
@Barker2022 ; @Hocquet2024]. We actively support community formation
by online training resources and communication channels. Together with the
interactive applications, these resources provide accessible entry points for
prospective developers and users.

**The OMA book.** _Orchestrating Microbiome Analysis with Bioconductor
(OMA)_ (https://microbiome.github.io/OMA/docs/devel/) is an online book that provides
a comprehensive introduction to the microbiome data science framework
in Bioconductor.
It offers practical guidance, reproducible workflows,
and training material for integrative microbiome analyses. The book is
versioned to match Bioconductor's biannual software release cycles. The work
incorporates rich feedback from the user community and microbiome training
courses. It complements the broader trend, where specific Bioconductor
data structures [@huber2015; @ramos2017; @huang2021] have been adopted as a
solution for multi-table data integration tasks in various
domains such as single-cell sequencing (OSCA book) [@amezquita2020]
and transciptomics [OSTA book](https://lmweber.org/OSTA/). These resources
demonstrate how standard data containers can support statistical
programming on interlinked multi-table data sets in a highly optimized way.

**Interactive applications.** Graphical user interfaces provide
accessible entry points to the framework. The iSEEtree and miaDash
packages provide interactive analysis and visualization tools for
microbiome data [@benedetti2025] (@fig-downstream_analyses). Their
interface supports automatic generation of source code and replicable
workflows as well as analysis templates for users without a strong
programming expertise. Moreover, the tidy R paradigm, available
through tidyomics, simplifies the syntax and can streamline the
workflows [@hutchison2024]. These tools allow users to perform basic
analyses and visualize data with ease.

**Community.** Its widespread adoption has made R one of the most cited software
resources of our time [@Pearson2025] and the introduction of the
shared data structures facilitates the broad interoperability of the
methods ecosystem. Bioconductor has a strong tradition in fostering the
R user community through shared data structures and training activities
[@drnevich2025]. An active community of developers and users thrives online,
supporting the software sustainability and promoting good practices in
scientific computing [@Wilson2017]. The community interacts via several online
communication channels, including Bioconductor Zulip, GitHub platform,
and an email list. The development of the framework is influenced by
feedback from the user community through these channels, regular
meetings and training workshops.

# Outlook {#sec-discussion}

```{r}
#| label: bioc_views
#| eval: true
#| echo: false
#| message: false

if( !require("BiocPkgTools") ){
    BiocManager::install("BiocPkgTools")
    library("BiocPkgTools")
}
# Get packages and info based on their BiocViews
pkgs <- biocPkgList()
#
fields <- sapply(pkgs[["biocViews"]], function(x){
    field <- "Other"
    if( any(grepl("genomic|genetic|gene|genom|DNA", x, ignore.case = TRUE)) ){
        field <- "Genomics"
    }
    if( any(grepl("proteomics|protein", x, ignore.case = TRUE)) ){
        field <- "Proteomics"
    }
    if( any(grepl("metabolomics|metabolome|metabolite|Lipidom|massspectro", x, ignore.case = TRUE)) ){
        field <- "Metabolomics"
    }
    if( any(grepl("transcripto|RNA-seq|RNA", x, ignore.case = TRUE)) ){
        field <- "Transcriptomics"
    }
    if( any(grepl("immuno", x, ignore.case = TRUE)) ){
        field <- "Immunology"
    }
    if( any(grepl("cytom", x, ignore.case = TRUE)) ){
        field <- "Cytometrics"
    }
    if( any(grepl("microarray|chip", x, ignore.case = TRUE)) ){
        field <- "Microarray"
    }
    if( any(grepl("single-cell|singlecell", x, ignore.case = TRUE)) ){
        field <- "Single-cell"
    }
    if( any(grepl("metagenom|microbiome|16S|microbiota|amplicon|shotgun|microb|metatranscript|metametabolo|metaproteo", x, ignore.case = TRUE)) ){
        field <- "Microbiome"
    }
    return(field)
})
pkgs[["Field"]] <- fields

pkg_stats <- pkgDownloadRank("mia")
pkg_stats <- paste0(pkg_stats[[1]], "%")
```

**The Bioconductor community** has a well-established reputation in
developing open, peer-reviewed software for life sciences. Its
strengths include the enforcement of minimal documentation standards,
semantic versioning, portability across operating platforms, quality
control (*e.g.* unit testing, continuous integration), usability (APIs),
and robustness (deprecation mechanisms), and a biannual release
cycle. During the past two decades Bioconductor has grown into a
repository of 2,300 actively maintained software packages, of which
`r sum(pkgs$Field == "Microbiome")` are classified under the Microbiome
Bioc task view. In order to enjoy the benefits of automation while
maintaining flexibility, data scientists routinely remix open-source
software into custom workflows. However, the lack of commonly adopted
data representation standards often hinders the development, sharing,
and application of statistical workflows. At the core of the
Bioconductor's data science strategy is the idea of shared data
representation. The enhanced interoperability can help to minimize
redundant efforts and facilitate the translation of methods between
different research areas [@DElia2023; @Barker2022] and, importantly,
facilitates the formation of broad developer and user communities.

**Microbiome research** has shifted towards integrative analyses
aiming at mechanistic and causal inference [@joos2024;
@bastiaanssen_bugs1_2023]. Technological advances in long-read
sequencing, functional assays, spatial profiling, and single-cell have
created an increasing demand for integrative data analysis
techniques. Thus, microbiome data is increasingly linked with other
omics along varying spatial and temporal dimensions. Our framework
addresses this need, together with parallel developments in other
areas of omics.

Recent developments in microbiome and multi-omics workflows increasingly
rely on Bayesian modeling and Gaussian processes to accommodate uncertainty
and temporal or spatial structures [@cheng2019]. Our framework is compatible
with these approaches, providing the structured input necessary for
probabilistic inference and scalable computation. Building on these
statistical foundations, the rapid emergence of foundation models
and deep learning approaches in biomedicine calls for robust, scalable
data representations [@awais2025]. Our framework, through its use of
standardized containers and modular workflows, is well-positioned to support
such advances. Applications include transfer learning across studies,
interpretable embeddings of microbiome multi-omic profiles,
computational prediction of microbial function and metabolites
[@mallick2019; @weiss2016], and integration with AI agents for hypothesis
generation and automated analysis, among others [@zou2025; @gao2024].

**Open data science infrastructures** have become critical components
of reproducible research and a vast body of applications critically
rely on open-source methods and workflows [@berg2020; @Barker2022;
@Hocquet2024]. Standardized data structures can enhance the transparency,
reliability, and explainability of analysis workflows. We have
documented the interoperable framework built on widely adopted data
containers, open data resources, analysis methods and reproducible
workflows for multi-modal microbiome research. Further developments could
include extended methodology for temporal, spatial, and multi-omic
analyses. Supporting interoperability between R and other
programming languages can facilitate the application of other ML
frameworks. As the field evolves, Bioconductor will continue
to integrate new tools and approaches to improve integrative analyses of
microbiome data.

# References
<!-- the pattern below controls the placement of the references -->
:::{#refs}
:::

## Consortia {#sec-consortium}

OMA consortium: Bastiaanssen, Thomaz; Beber, Moritz E.; Bektanov, Aituar;
Benedetti, Giulio; Borman, Tuomas; Braccia, Domenick J.; Calgaro, Matteo; Callan,
Danielle; Cao, Yang; Chaudhary, Jiya; Corrada Bravo, Héctor;

Courbayre Dussau, Basil; Das, Sneha; Eckermann, Henrik; Erawijantari, Pande;
Ernst, Felix G.M.; Gaudron-Parry, Elliot; Gopalakrishnan, Shyam; Hakula, Maija;
Hanhineva, Kati; Havulinna, Aki; Hindström, Rasmus; Huang, Ruizhu; Jeba, Akewak;
Lahti, Leo; Limborg, Morten; Lindgren, Himmi; Liu, Yihan; Mallick, Himel;

Mongad, Dattatray; Müller, Christian L.; Muluh, Geraldson; Nynäs, Signe;
Pasanen, Jesse; Paulo, Joao; Pelto, Juho; Peschel, Stefanie; Potbhare, Renuka;
Pralas, Théotime; Ramos, Marcel; Raulo, Aura; Sannikov, Artur; Seraidarian, Ely;
Serizay, Jacques; Shetty, Sudarshan; Shigdel, Rajesh; Suksi, Vilhelm;
Tammi, Eetu; Valderamma, Benjamin; Waldron, Levi; Yang, Lu.

## Contributions {#sec-contributions}

LL and TB initiated and coordinated the work. All authors contributed
material, participated in manuscript preparation, and approved the final
version.

## Acknowledgments {#sec-acknowledgments}

This project received funding from the European Union’s Horizon 2020
research and innovation programme under grant agreement No 952914. TB
was additionally supported by the University of Turku Graduate School.

We would like to thank the following individuals for their support for
the project: Ashma, Richa; Bayal, Nitin; Benchraka, Chouaib; Dagnaud, Axel;
Dallier, Teo; de Gunst, Noah; Faust, Karoline; Finn, Rob; Hillman, Samuel;
Pagés, Hervé; Richardson, Lorna; Rios Garza, Daniel;
Robinson, Mark; Ruuskanen, Matti; Salehi, Mahkameh.

We also extend our thanks to all contributors of the open-source resources used
in this work.

The authors wish to acknowledge CSC – IT Center for Science, Finland,
for computational resources.
